Tree species and structural diversity are important determinants of ecosystem function across disturbed southern African woodlands

### 2020-09-17 manuscript/regional_befr_manuscript.tex
Submitted

\title{Tree biodiversity and ecosystem function relationship across southern African woodlands}
\author{}
\date{}

@@

{\LARGE{Title: Tree species and structural diversity are important determinants of ecosystem function across southern African woodlands}}

\vspace{1cm}

Authors: John L. Godlee\textsuperscript{1}, Casey M. Ryan\textsuperscript{1}, David Bauman\textsuperscript{2,3}, Samuel J. Bowers\textsuperscript{1}, Joao M. B. Carreiras\textsuperscript{4}, Antonio Valter Chisingui\textsuperscript{5}, Joris P. G. M. Cromsigt\textsuperscript{6,7,8}, Dave J. Druce\textsuperscript{9,10}, Manfred Finkch\textsuperscript{11}, Francisco Maiato Gon\c{c}alves\textsuperscript{5}, Ricardo M. Holdo\textsuperscript{12}, Steve Makungwa\textsuperscript{13}, Iain M. McNicol\textsuperscript{1}, Edward T. A. Mitchard\textsuperscript{1}, Anderson Muchawona\textsuperscript{14}, Rasmus Revermann\textsuperscript{11,15}, Natasha Sofia Ribeiro\textsuperscript{16}, Abel Siampale\textsuperscript{17}, Stephen Syampungani\textsuperscript{18}, Jos\'{e} Jo\~{a}o Tchamba\textsuperscript{5}, Hemant G. Tripathi\textsuperscript{19}, Johannes Wallenfang\textsuperscript{11}, Mariska te Beest\textsuperscript{20,21,22}, Mathew Williams\textsuperscript{1}, Kyle G. Dexter\textsuperscript{1,23}

1: School of GeoSciences, University of Edinburgh, Edinburgh, EH9 3FF, United Kingdom \\
2: Environmental Change Institute, School of Geography and the Environment, University of Oxford, Oxford, United Kingdom \\
3: Laboratoire d'\'{E}cologie V\'{e}g\'{e}tale et Biogéochimie, CP 244, Universit\'{e} Libre de Bruxelles, Brussels, Belgium \\
4: National Centre for Earth Observation (NCEO), University of Sheffield, Hicks Building, Hounsfield Road, Sheffield, S3 7RH, United Kingdom \\
5: Herbarium of Lubango, ISCED Hu\'{i}la, Sarmento Rodrigues Str. No. 2, CP. 230, Lubango, Angola \\
6: Department of Wildlife, Fish, and Environmental Studies, Swedish University of Agricultural Sciences, Ume\aa, Sweden \\
7: Department of Zoology, Centre for African Conservation Ecology, Nelson Mandela University, Port Elizabeth, South Africa \\
8: Copernicus Institute of Sustainable Development, Utrecht University, 3584CS Utrecht, the Netherlands \\
9: Ecological Advice, Ezemvelo KZN Wildlife, Hluhluwe-iMfolozi Park, South Africa \\
10: School of Life Sciences, University of KwaZulu-Natal, Pietermaritzburg, South Africa \\
11: Biodiversity, Evolution and Ecology of Plants, Institute of Plant Sciences and Microbiology, University of Hamburg, Ohnhorststr. 18, 22609 Hamburg, Germany \\
12: Odum School of Ecology, University of Georgia, 140 E. Green St., Athens, GA 30602, USA \\
13: Lilongwe University of Agriculture and Natural Resources (LUANAR), Lilongwe, Malawi \\
14: Forest Research Center, 1 Orange Groove Highlands, Harare, Zimbabwe \\
15: Faculty of Natural Resources and Spatial Sciences, Namibia University of Science and Technology, Windhoek, Namibia \\
16: Department of Forest Engineering, Faculty of Agronomy and Forest Engineering, UEM Av. Julius Nyerere, 3453, Campus Universitario, Maputo, Mozambique \\
17: Forestry Department Headquarters - MLNR, Cairo Road, Lusaka, Zambia \\
18: School of Natural Resources, Copperbelt University, Kitwe, Zambia \\
19: Faculty of Biological Sciences, University of Leeds, LS2 9JT, United Kingdom\\
20: Copernicus Institute of Sustainable Development, Utrecht University, 3508 TC Utrecht, The Netherlands \\
21: Centre for African Conservation Ecology, Nelson Mandela University, Port Elizabeth, 6031, South Africa \\
22: South African Environmental Observation Network, Grasslands-Forests-Wetlands Node, Montrose, 3201, South Africa \\
23: Royal Botanic Garden Edinburgh, Edinburgh, EH3 5LR, United Kingdom \\

\vspace{1em}
Corresponding author:

John L. Godlee

johngodlee@gmail.com

School of GeoSciences, University of Edinburgh, Edinburgh, EH9 3FF, United Kingdom

\section{Acknowledgements}

This work is funded by a NERC E3 Doctoral Training Partnership PhD studentship at the University of Edinburgh (John L. Godlee, Grant No. NE/L002558/1). The data for this study was contributed by a number of independently funded projects and was assembled and prepared by SEOSAW (A Socio-Ecological Observatory for Southern African Woodlands, \url{https://seosaw.github.io}), an activity of the Miombo Network and a NERC-funded project (Grant No. NE/P008755/1). Revisions of the SEOSAW dataset were funded by SavannaChange, a GCRF/University of Edinburgh funded project. We thank all data providers and the field assistance they received when collecting plot data. JMBC was supported by the Natural Environment Research Council (Agreement PR140015 between NERC and the National Centre for Earth Observation).

\section{Conflict of interest statement}

All authors declare no conflict of interest regarding this study.

\section{Biosketch}

SEOSAW (A Socio-Ecological Observatory for Southern African Woodlands, \url{https://seosaw.github.io}) aims to understand the response of southern African woodlands to global change. The goal of SEOSAW is to produce novel analyses of the determinants of ecosystem structure and function for the southern Africa region, based on syntheses of plot data. Additionally the group hopes to develop infrastructure for a long-term regional plan for plot remeasurement in the southern African region. While working on a multitude of diverse projects in the dry tropics at large, all authors have a broad interest in community ecology and ecosystem assemblage in southern African woodlands.

@@

\LARGE{Title: Tree species and structural diversity are important determinants of ecosystem function across southern African woodlands}

@@

\section{Abstract}

\textbf{Aim:} Positive correlations between tree species diversity and ecosystem function have been widely documented, but the nature of the relationship in southern African savanna/woodlands, which experience high levels of disturbance through fire and ecophysiological stress, is less clear. It is posited that high levels of disturbance moderate the effects of competition in savannas, weakening the correlation between biodiversity and niche complementarity that drives ecosystem function. Here, we explore the relationship between tree species diversity and aboveground biomass across southern African savannas and woodlands, while controlling for variation in stem density, resource availability, disturbance through fire, and vegetation types to build a general understanding of the biodiversity - ecosystem function relationship in this understudied ecological context.

\textbf{Location:} Southern African savannas and woodlands
\textbf{Time period:} 2010-2019

@@

\textbf{Methods:} We used a network of \nplots{} savanna/woodland tree plots located across the southern African sub-continent. We used Structural Equation Modelling with path analysis to determine the relationship between tree species diversity and aboveground woody biomass, while accounting for the interactive effects of resource availability, disturbance by fire, stem density and vegetation type.

\textbf{Results:} We found a positive effect of tree species diversity on aboveground biomass, across vegetation types, observed mainly via the increasing effect of woodland structural diversity. We also found that the effect of tree species diversity on biomass increases with stem density, with a minimum threshold of \textapprox{}180 stems ha\textsuperscript{-1}. Finally, we found that resource availability affects biomass in southern African woodlands mainly indirectly, via its effect on species diversity, suggesting that tree diversity may have been under-appreciated as a determinant of savanna structure.

@@

>>>>

\title{Tree species and structural diversity are important determinants of ecosystem function across disturbed southern African woodlands}
\author{}
\date{}

@@

@@

\LARGE{Title: Tree species and structural diversity are important determinants of ecosystem function across disturbed southern African woodlands}

@@

\section*{Abstract}

\textbf{Aim:} Positive correlations between tree species diversity and ecosystem function have been widely documented, but the nature of the relationship in southern African savanna/woodlands, which experience high levels of disturbance through fire and ecophysiological stress, is less clear. It is posited that disturbance moderates the effects of competition in savannas, weakening the correlation between biodiversity and niche complementarity that drives ecosystem function. We explore the relationship between tree species diversity and biomass across southern African woodlands, accounting for variation in stem density, resource availability, disturbance, and vegetation type to build general understanding of the biodiversity-ecosystem function relationship in this understudied ecological context.

\textbf{Location:} Southern African savannas and woodlands

\textbf{Time period:} 2010-2019

@@

\textbf{Methods:} We used a network of \nplots{} woodland tree plots located across southern Africa. We used Structural Equation Modelling to determine the relationship between tree species diversity and aboveground woody biomass, while accounting for the interactive effects of resource availability, disturbance by fire, stem density and vegetation type.

\textbf{Results:} We found positive effects of tree species diversity on aboveground biomass, across vegetation types, observed mainly indirectly via the increasing effect of woodland structural diversity. We also found that the effect of tree species diversity on biomass increases with stem density, with a minimum threshold of \textapprox{}180 stems ha\textsuperscript{-1}. Finally, we found that resource availability affects biomass in southern African woodlands mainly indirectly, via species diversity, suggesting that tree diversity may have been under-appreciated as a determinant of savanna structure.

@@

\section{Data accessibility statement}

The data that support the findings of this study are available on request from the corresponding author. The data are not publicly available at the time of submission due to privacy restrictions regarding plot locations. An anonymised version will be made available in a data repository following review.
